---
title: Introducing `geom_terrorbar()`
date: 2023-01-01T09:12:12+01:00
background: intro.jpg
tags:
    - R
    - ggplot
    - Open Science
featured: true
# draft: true
# for Quarto posts (must rename to .qmd)
format: hugo-md
# jupyter: python3
---

<style>
    .serif {
        font-family: "Times New Roman", serif
    }
</style>

## Motivation

I was reading [Adam Mastroianni](https://twitter.com/a_m_mastroianni)'s excellent blog post on [peer review](https://experimentalhistory.substack.com/p/the-rise-and-fall-of-peer-review) (it's really good, go check it out) because it's Christmas time and I don't have to be a code monkey for a while. To give an example of the kind of rubbish that somehow makes it through peer review, Adam calls to his aid a tweet many of you will have seen but for me it was a revelation:

{{< tweet 1597355324008108034 >}}

Going by the lively discussion in the thread, many people seem just as appreciative of this galaxy-brain-level of problem solving as I was.

Now, I'd heard of Advances in Materials Science and Engineering, the journal that published the [now retracted paper](https://www.hindawi.com/journals/amse/2022/3802603/) that this figure is from, on the [Everything Herz podcast](https://everythinghertz.com/) hosted by the young researcher of the year [Dan Quintana](https://twitter.com/dsquintana) and the curmudgeonly self-indulgent academia survivor [James Heathers](https://twitter.com/jamesheathers) (again, go check it out!) but I wasn't ready for quite this level of hilarity.

I simply think that having <span class="serif">T</span>s for errorbars is wonderful and everyone should start doing it from now on. That's why I bring you today `ggterror`, an extension to `R`'s `ggplot2` package that allows you to do exactly that. Because it's Christmas time and though I don't have to be a code monkey for a while, deep down I guess I know that I really am a code monkey...

Unlike the graphics in the original -- and this word is doing a lot of work here -- paper, this extension produces actual errorbars.

## Installation

`ggterror` is hosted on my GitHub so help yourself:

```{r}
#| eval: false
remotes::install_github("mivalek/ggterror")
```

## Usage

The package contains a single function `geom_terrorbar()` that should be used in much the same way as `ggplot2::geom_errorbar()`

```{r}
#| fig-height: 4
#| fig-width: 5
#| label: peng-fig
#| fig-cap:  _Figure 1_. Mean body weight per penguin species. Terrorbars represent +/- 2 SE of the mean.


library(ggplot2)
library(ggterror)

palmerpenguins::penguins |>
    dplyr::filter(!is.na(body_mass_g)) |>
    dplyr::group_by(species) |> 
    dplyr::summarise(
        mean = mean(body_mass_g),
        se = sd(body_mass_g) / sqrt(dplyr::n())
        ) |>
        ggplot(aes(x = species, y = mean)) + geom_terrorbar(aes(ymin = mean - 2 * se, ymax = mean + 2 * se)) +
        geom_point() +
        theme_bw()

```

On top of the usual aesthetics you'd pass to `ggplot2::geom_errorbar()`, this function takes a few additional arguments that govern the overal look of the <span class="serif">T</span>s and the connecting lines:

- `connect` - logical. Whether or not the <span class="serif">T</span>s should be connected with a line.
- `fontface` - one of `c("plain", "bold", "italic")`
- `size` - size of the <span class="serif">T</span>s in pt units
- `line_type` - type of connecting line when `connect = TRUE`
- `line_width` - width of connecting line when `connect = TRUE`
- `line_alpha` - alpha for connecting line when `connect = TRUE`. By default, `line_alpha = 0.6 * alpha` of the <span class="serif">T</span>s

Hope you have fun creating some true classics!

## Bonus: _that_ plot

As a bonus ('cause it's Christmas time), I thought I'd recreate _that_ plot:

```{r}
#| label: that-fig
#| fig-cap: _Figure 2_. That figure, recreated. Terrorbars represent fuck all.
#| warning: false
#| echo: true
#| fig-height: 4
#| fig-width: 5

library(ggtext)

df <- tibble::tibble(x = factor(1:5, labels=c("blank", "pink", "lysozyme", "hOGG1", "Endo IV")), y = c(3.5, .5, 3.4, 3.2, 3), ymax = y + .2,ymin = y - .2, size = 15)

df |> ggplot(aes(x, y)) +
  geom_terrorbar(aes(ymax = ymax, ymin = ymin), connect = FALSE) +
  geom_bar(stat = "identity", aes(fill = x), colour="black", width = .65, lwd = .7) +
  labs(x = "", y = "Current (<em>&mu;</em>A)") +
  scale_y_continuous(limits = c(0,4.3), expand=c(0,0)) +
  theme_classic() +
  scale_fill_manual(values=c("#f3c08f", "#f9ee87", "#3eaba8", "#8fc4f1", "#cecece")) +
  theme(
    panel.background = element_rect(colour = "black", fill = NA, size = .85),
    axis.text = element_text(family="serif", size = 15),
    axis.title.y = element_markdown(family = "serif", size = 15),
    axis.ticks = element_blank()) +
  guides(fill = "none")
```

So many wonderful design choices! Chef's kiss, mwah...

Have a happy new year, all y'all!
